---
title: "Quantium Virtual Internship - Retail Strategy and Analytics - Task 1"
mainfont: Roboto
monofont: Consolas
output:
 pdf_document:
 df_print: default
 highlight: tango
 keep_tex: yes
 latex_engine: xelatex
header-includes:
 \usepackage{fvextra}

\DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
# set options for R markdown knitting
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(linewidth=80)
```


```{r knitr line wrap setup, include=FALSE}
# set up line wrapping in MD knit output
library(knitr)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options)
{
 # this hook is used only when the linewidth option is not NULL
 if (!is.null(n <- options$linewidth))
 {
 x = knitr:::split_lines(x)
 # any lines wider than n should be wrapped
 if (any(nchar(x) > n))
 x = strwrap(x, width = n)
 x = paste(x, collapse = "\n")
 }
 hook_output(x, options)
})
```


```{r 0 Load Libraries, results = 'hide'}
#### Example code to install packages
#install.packages("data.table")
#### Load required libraries
library(data.table)
library(ggplot2)
library(ggmosaic)
library(readr)

filePath <- "D:/DOC/"
transactionData <- fread(paste0(filePath,"QVI_transaction_data.csv"))
customerData <- fread(paste0(filePath,"QVI_purchase_behaviour.csv"))
```

## Exploratory data analysis
### Examining transaction data
```{r Examining transaction data}
#### Examine transaction data
str(transactionData)
head(transactionData)
```

```{r Convert DATE to date format}
#### Convert DATE column to a date format
#### A quick search online tells us that CSV and Excel integer dates begin on 30 Dec 1899
transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30")
```


```{r Summary of PROD_NAME}
#### Examine PROD_NAME
str(transactionData$PROD_NAME)
head(transactionData$PROD_NAME)
```

```{r Further examine PROD_NAME}
#### Examine the words in PROD_NAME to see if there are any incorrect entries
#### such as products that are not chips
productWords <- data.table(unlist(strsplit(unique(transactionData[, PROD_NAME]), " ")))
setnames(productWords, 'words')
```

```{r}
# Remove digits, and special characters, and then sort the distinct words by frequency of occurrence.
#### Removing digits
productWords <- productWords[grepl("\\d", words) == FALSE, ]

#### Removing special characters
productWords <- productWords[grepl("[:alpha:]" , words), ]

#### Let's look at the most common words by counting the number of times a word appears and
#### sorting them by this frequency in order of highest to lowest frequency
productWords[, .N, words][order(N, decreasing = TRUE)]
```

```{r}
#### Remove salsa products
transactionData[, SALSA := grepl("salsa", tolower(PROD_NAME))]
transactionData <- transactionData[SALSA == FALSE, ][, SALSA := NULL]
```

```{r initial summary}
#### Summarise the data to check for nulls and possible outliers
summary(transactionData)
```
There are no nulls in the columns but product quantity appears to have an outlier
which we should investigate further. Let's investigate further the case where 200
packets of chips are bought in one transaction.

```{r }
#### Filter the dataset to find the outlier
filter(transactionData, transactionData$PROD_QTY == 200)
```
There are two transactions where 200 packets of chips are bought in one transaction
and both of these transactions were by the same customer. 
```{r}
#### Let's see if the customer has had other transactions
filter(transactionData, transactionData$LYLTY_CARD_NBR == 226000)
```

It looks like this customer has only had the two transactions over the year and is
not an ordinary retail customer. The customer might be buying chips for commercial
purposes instead. We'll remove this loyalty card number from further analysis.

```{r}
#### Filter out the customer based on the loyalty card number
transactionData <- filter(transactionData, transactionData$LYLTY_CARD_NBR != 226000)

#### Re-examine transaction data
summary(transactionData)
```

That's better. Now, let's look at the number of transaction lines over time to see
if there are any obvious data issues such as missing data.

```{r}
#### Count the number of transactions by date
transactions_by_day <- transactionData[, .N, DATE][order(DATE)]
transactions_by_day
```

There's only 364 rows, meaning only 364 dates which indicates a missing date. Let's
create a sequence of dates from 1 Jul 2018 to 30 Jun 2019 and use this to create a
chart of number of transactions over time to find the missing date.

```{r fig.align = "center"}
#### Create a sequence of dates and join this the count of transactions by date
# create a column of dates that includes every day from 1 Jul 2018 to 30 Jun 2019, and join it onto the data to fill in the missing day.
new_date <- data.frame(DATE=as.Date(seq(as.Date("2018-07-01"), as.Date("2019-06-30"), "day")))
transactions_by_day <- merge(x=new_date,y=transactions_by_day,by="DATE",all.x=TRUE)
#### Setting plot themes to format graphs
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))
#### Plot transactions over time
ggplot(transactions_by_day, aes(x = DATE, y = N)) +
 geom_line() +
 labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
 scale_x_date(breaks = "1 month") +
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```
We can see that there is an increase in purchases in December and a break in late
December. Let's zoom in on this.

```{r fig.align = "center"}
dec <- filter(transactions_by_day, month(transactions_by_day$DATE) == 12)
#### Setting plot themes to format graphs
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))
#### Plot transactions over time
ggplot(dec, aes(x = DATE, y = N)) +
 geom_line() +
 labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
 scale_x_date(breaks = "1 day") +
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```
We can see that the increase in sales occurs in the lead-up to Christmas and that
there are zero sales on Christmas day itself. This is due to shops being closed on
Christmas day.
Now that we are satisfied that the data no longer has outliers, we can move on to
creating other features such as brand of chips or pack size from PROD_NAME. We will
start with pack size.

```{r Create pack size}
#### Pack size
#### We can work this out by taking the digits that are in PROD_NAME
transactionData[, PACK_SIZE := parse_number(PROD_NAME)]
#### Always check your output
#### Let's check if the pack sizes look sensible
transactionData[, .N, PACK_SIZE][order(PACK_SIZE)]
```

The largest size is 380g and the smallest size is 70g - seems sensible!

```{r }
#### Let's plot a histogram of PACK_SIZE since we know that it is a categorical variable and not a continuous variable even though it is numeric.
ggplot(transactionData, aes(x=PACK_SIZE))+
  geom_histogram(colour="black", fill="white", bins = 30)
```
Pack sizes created look reasonable.
Now to create brands, we can use the first word in PROD_NAME to work out the brand
name...
```{r Create brand name}
#### Brands
# Create a column which contains the brand of the product, by extracting it from the product name.
transactionData[, BRAND := toupper(substr(PROD_NAME, 1, regexpr(pattern = ' ', PROD_NAME) -1))]
#### Checking brands
transactionData[, .N, by = BRAND][order(-N)]
```

Some of the brand names look like they are of the same brands - such as RED and
RRD, which are both Red Rock Deli chips. Let's combine these together.

```{r Clean brand names}
#### Clean brand names
transactionData[BRAND == "RED", BRAND := "RRD"]
transactionData[BRAND == "SNBTS", BRAND := "SUNBITES"]
transactionData[BRAND == "INFZNS", BRAND := "INFUZIONS"]
transactionData[BRAND == "WW", BRAND := "WOOLWORTHS"]
transactionData[BRAND == "SMITH", BRAND := "SMITHS"]
transactionData[BRAND == "NCC", BRAND := "NATURAL"]
transactionData[BRAND == "DORITO", BRAND := "DORITOS"]
transactionData[BRAND == "GRAIN", BRAND := "GRNWVES"]

#### Check again
transactionData[, .N, BRAND][order(N)]
```

### Examining customer data
Now that we are happy with the transaction dataset, let's have a look at the
customer dataset.

```{r 1 Exploratory data analysis}
#### Examining customer data
head(customerData)
summary(customerData)
customerData[, .N, LIFESTAGE][order(-N)]
customerData[, .N, PREMIUM_CUSTOMER][order(-N)]
```

```{r }
#### Merge transaction data to customer data
data <- merge(transactionData, customerData, all.x = TRUE)
```

As the number of rows in `data` is the same as that of `transactionData`, we can be
sure that no duplicates were created. This is because we created `data` by setting
`all.x = TRUE` (in other words, a left join) which means take all the rows in
`transactionData` and find rows with matching values in shared columns and then
joining the details in these rows to the `x` or the first mentioned table.

Let's also check if some customers were not matched on by checking for nulls.
```{r Check for missing customer details}
lapply(data, function(x)sum(is.na(x)))
```

Great, there are no nulls! So all our customers in the transaction data has been
accounted for in the customer dataset.

```{r Code to save dataset as a csv}
fwrite(data, paste0(filePath,"QVI_data.csv"))
```

## Data analysis on customer segments 
Let's start with calculating total sales by LIFESTAGE and PREMIUM_CUSTOMER and
plotting the split by these segments to describe which customer segment contribute
most to chip sales.

```{r fig.width = 10, fig.align = "center"}
#### Total sales by LIFESTAGE and PREMIUM_CUSTOMER
# Over to you! Calculate the summary of sales by those dimensions and create a plot.
```